Structuring Theories with Implicit Morphisms

### 2018-10-14 appl.tex
update

\ednote{@DM: add example, e.g., a definition of Monoid that uses different names}

\paragraph{Definitional Extensions}
We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
\begin{compactitem}
 \item only constant declarations with definiens,
 \item only include declarations of theories that are definitional extensions of $S$.
\end{compactitem}
\ednote{@DM: give example that extends Group with the theorem $(x^{-1})^{-1}\doteq x$ (with omitted proof)}

@@

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
\end{compactitem}

\ednote{@DM: repeat previous example now with \keyword{definitional} and give another definitional extensions of Group; this new extension may already use the implicitly available definition of the first extension}

@@

\ednote{@DM: add the isomorphism of DG2G}
\begin{example}\label{group:iso}
We give the morphism $\cn{G2DG}$.

\end{example}

While the first one of these declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphism.
Otherwise, the uniqueness condition would be violated.
Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $c:E\in\flt{A}$.
Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.

>>>>

\begin{example}
	Consider a variation of the theory $\cn{Monoid}$ from \autoref{syn:incl} in a different library:
	\begin{mmtcode}
Monoid2 =
  M: type
  connective : M ⟶ M ⟶ M	 # 1 ∘ 2 
  neutral	: M
\end{mmtcode}
This theory is isomorphic to the previously introduced theory $\cn{Monoid}$ under the trivial renaming
	\begin{mmtcode}
Monoids : Monoid2 -> Monoid =
  M 		:= U
  connective 	:= op 
  neutral	:= unit
\end{mmtcode}
\end{example}

\paragraph{Definitional Extensions}
We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
\begin{compactitem}
 \item only constant declarations with definiens,
 \item only include declarations of theories that are definitional extensions of $S$.
\end{compactitem}
\begin{example}
	Imagine extending the theory $\cn{Group}$ from \autoref{syn:incl} by provable theorems:
		\begin{mmtcode}
Group_Theorems =
  include Group
  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)
\end{mmtcode}
\end{example}

@@

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
\end{compactitem}
\begin{example}
	We adapt our theory $\cn{Group\_Theorems}$ with the newly introduced keyword, which allows us to extend $\cn{Group}$ directly while having the new theorem $\cn{inverse\_idem}$ available:
		\begin{mmtcode}
Group_Theorems =
  definitional include Group
  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)

More_Theorems =
  include ?Group
  some_other_theorem : ⊦ ??	= (proof using inverse_idem)
\end{mmtcode}
\end{example}

@@

\begin{example}\label{group:iso}
Having declared the morphism $\cn{DG2G}$ (as in \autoref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
\begin{mmtcode}
implicit G2DG : Group -> DivisionGroup =
  U := U
  op := [a,b] a / ((unit/unit) / b) 
  unit := unit/unit
  inverse := [a] (unit/unit) / a
\end{mmtcode}
\end{example}

While the first one of these isomorphism declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphisms.
Otherwise, the uniqueness condition would be violated.
Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $c:E\in\flt{A}$.
Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.

### 2018-10-14 example.tex
update

A common problem when definition modular theory hierarchies is that the most natural include-hierarchy for the most important theories is not necessarily the same as the most comprehensive hierarchy.
For example, Ex.~\ref{syn:incl} defines \cn{Group} with an include from \cn{Monoid}.
Instead, we could include \cn{Monoid} into the the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
This change is not possible in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents an insurmountable refactoring problem.
So instead we could systematically build a hierarchy that uses every intermediate theory (as done in \cite{mathscheme}).
But this yields a very deep and complex hierarchy that is hard to navigate for casual users.
Moreover, it does not protect us from later on discovering yet another intermediate theory that should have been added.

@@

\begin{figure}\begin{center}
\includegraphics[width=0.6\textwidth]{bands}\end{center}
\caption{The Lattice of Varieties of Bands}\label{fig:bands}
\end{figure}

>>>>

A common problem when defining modular theory hierarchies is that the most natural include-hierarchy for the most important theories is not necessarily the same as the most comprehensive hierarchy.
For example, Ex.~\ref{syn:incl} defines \cn{Group} with an include from \cn{Monoid}.
Instead, we could include \cn{Monoid} into the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
This change is not possible in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents an insurmountable refactoring problem.
So instead we could systematically build a hierarchy that uses every intermediate theory (as done in \cite{mathscheme}).
But this yields a very deep and complex hierarchy that is hard to navigate for casual users.
Moreover, it does not protect us from later on discovering yet another intermediate theory that should have been added.

@@

\begin{figure}\begin{center}
\begin{tikzpicture}[scale=1.5]
  \node (s1) at (0,0) {$zxyz=zxzyz$};

  \node (s2a) at (-1,-1) {$zxy=zxzy$};
  \node (s2b) at (1,-1) {$yxz=yzxz$};
  \draw (s1) to (s2a) {};
  \draw (s1) to (s2b) {};

  \node (s3a) at (-2,-2) {$xy=xyx$};
  \node (s3b) at (0,-2) {$zxyz=zyxz$};
  \node (s3c) at (2,-2) {$xy=yxy$};
  \draw (s2a) to (s3a) {};
  \draw (s2a) to (s3b) {};
  \draw (s2b) to (s3b) {};
  \draw (s2b) to (s3c) {};

  \node (s4a) at (-1.5,-3) {$zxy=zyx$};
  \node (s4b) at (0,-3) {$xyx=x$};
  \node (s4c) at (1.5,-3) {$xyz=yxz$};
  \draw (s3a) to (s4a) {};
  \draw (s3b) to (s4a) {};
  \draw (s3b) to (s4b) {};
  \draw (s3b) to (s4c) {};
  \draw (s3c) to (s4c) {};

  \node (s5a) at (-1.5,-4) {$xy=x$};
  \node (s5b) at (0,-4) {$xy=yx$};
  \node (s5c) at (1.5,-4) {$xy=y$};
  \draw (s4a) to (s5a) {};
  \draw (s4a) to (s5b) {};
  \draw (s4b) to (s5a) {};
  \draw (s4b) to (s5c) {};
  \draw (s4c) to (s5b) {};
  \draw (s4c) to (s5c) {};

  \node (s6) at (0,-5) {$x=y$};
  \draw (s5a) to (s6) {};
  \draw (s5b) to (s6) {};
  \draw (s5c) to (s6) {};

  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
  %\node at (t1.north west) {$S$};
  %\node at (t1) {$D$};
  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
  %\node at (t2.north west) {$T$};
  %\node at (t2) {$C$};
  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
\end{tikzpicture}

%\includegraphics[width=0.6\textwidth]{bands}
\end{center}
\caption{The Lattice of Varieties of Bands}\label{fig:bands}
\end{figure}

### 2018-10-14 mmt.tex
update

\paragraph{Logics in \mmt}
\ednote{explain that we use a fixed instance of MMT which adds LF and FOL; only relevant for examples}

@@

Moreover, we use a few constants from a type systems\ednote{finish}$\forall,\doteq$ etc.
\end{example}

>>>>

\paragraph{Logics in \mmt}
\ednote{explain that we use a fixed instance of MMT which adds LF and FOL; only relevant for examples}
\mmt itself is foundation independent and parmetric in a set of typing rules. We omit the details on implementing type systems and logical frameworks in \mmt, and will instead assume theories for the logical framework LF and an implementation of first-order logic for the examples in the remainder of this paper.

Specifically, we assume a theory $\cn{LF}$ that implements the dependently-typed lambda calculus LF. This theory provides symbols for dependent function types $\prod_{c:A}B(x)$ (in surface syntax: \lstinline|{x:A}B x|), simple function types $A\to B$ (as abbreviation for $\prod_{\_:A}B$, lambda abstractions $\lambda x:A.\; t(x)$ (in surface syntax: \lstinline|[x:A] t x|) and function application $f a$.

The theory \cn{LF} is used as a meta-theory for a theory \cn{FOL} implementing first-order logic with equality. This theory provides symbols for the quantifiers $\forall$ and $\exists$ (implemented via higher-order abstract syntax, i.e. the proposition $\forall x.\; P(x)$ is written as $\forall$\lstinline|[x:U]P x| for a fixed type $U$), equality $\doteq$ and the usual logical connectives $\neg,\wedge,\vee$, etc.

@@

Moreover, we use the constants from the theories $\cn{LF}$ and $\cn{FOL}$ described above.
\end{example}
